---
title: "Code to calculate the impact indicator"
author: "Mukhtar Yahaya, Sabrina Kumschick, Sandra MacFadyen, Pietro Landi, Cang Hui"
date: "`r Sys.Date()`"
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# impact.indicator

<!-- badges: start -->
<!-- badges: end -->


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Description

This R Markdown demonstrates the computation of an impact indicator for biological invasions using the `impact_indicator()`. The `impact_indicator()` feeds in species occurrence cube from the `b3gbi::process_cube()` using `taxaFun()` and processed 
Environmental Impact Classification of Alien Taxa (EICAT) impact score of species
using `impact_cat()`.

```{r load packages }
# Load packages
library(b3gbi)       # Biodiversity indicators for data cubes
library(tidyverse)   # Data wrangling and visualisation
library(sf)          # Spatial features


# Source functions
source("R/impact_indicator.R")
source("R/taxa_cube.R")
source("R/impact_cat.R")

```

```{r}
# load the shapefile for study region
countries_sf<-readRDS("Data/countries_shapefile.rds")
SA.sf<-filter(countries_sf,name=="South Africa") %>% select(name,geometry)
plot(SA.sf, main = "South African map")
```

# Process occurrence cube
```{r}
# load the GBIF occurrence data for taxa
taxa_Acacia<-readRDS("Data/taxa_Acacia.rds")
acacia_cube<-taxaFun(taxa=taxa_Acacia,
                    region=SA.sf,
                    res=0.25,
                    first_year=2010)

acacia_cube$cube
head(acacia_cube$coords)
```
# Aggregate impact scores for each species

There are often several impact records per species in different mechanisms and regions. There is need to aggregate these impact records for each species. The 
`impact_cat()` aggregates impact using ***max***, ***mean*** and  ***max_mech*** as metrics as used by different studies. 

- ***max***: maximum impact score across all records for the species\
- ***mean***: mean impact score across all records\
- ***max_mech***: sum of the maximum impact per mechanisms

```{r }
eicat_data<-readRDS("Data/eicat_data.rds")
full_species_list<-sort(unique(acacia_cube$cube$data$scientificName))
eicat_data %>% select(scientific_name,impact_region,impact_mechanism,
                      impact_category) %>% 
  head(10)
agg_impact<-impact_cat(impact_data=eicat_data,
                     species_list=full_species_list,
                     col_impact=NULL,
                     col_name=NULL)
agg_impact
```
# Compute impact risk map


The impact risk map shows the impact score for each site, where multiple species can be present. To compute the risk per site, aggregated scores across species at each site are needed. The 
`impact_indicator()` uses ***max***, ***sum*** and ***mean*** metrics to aggregate
impact scores as proposed by Boulesnane-Guengant et al., (in preparation). The combinations of aggregation metrics for each species and site leads to five type of
indicators, namely, ***precautionary***, ***precautionary cumulative***, ***mean***, ***mean cumulative*** and ***cumulative***.\


- ***precautionary***: maximum score across species' max in each site\
- ***precautionary cumulative***: cumulative score across species' max in each site\
- ***mean***: mean score across species' mean in each site\
- ***mean cumulative***: cumulative score across species' mean in each site\
- ***cumulative***: cumulative score across species' sum of maximum score per mechanism\
```{r}
impact_value<-impact_indicator(cube=acacia_cube$cube,
                               impact_data = eicat_data,
                               col_impact=NULL,
                               col_name=NULL,
                               type = "mean cumulative",
                               coords=acacia_cube$coords)
```
### impact risk map
```{r}
#impact risk map
#visualise last four years for readability
impact_value$sitedf%>% 
  gather(-c(siteID,X,Y),key="year",value="impact") %>% 
  na.omit() %>% 
  filter(year>=2021) %>% 
  ggplot() +
  geom_tile(
    aes(x=X,y=Y,fill=impact),color="black")+
  geom_sf(data = SA.sf, fill = NA, color = "black", alpha = 0.5)+
  scale_fill_gradient2(low = "forestgreen",
                       mid = "yellow",
                       high = "red")+
  theme_minimal() +
   labs(
    title = "impact risk map (mean cumulative)",
    y = "Latitude", x="Longitude"
  )+
  theme(text=element_text(size=14))+
  facet_wrap(~year)
```

# Compute impact indicators

To compute the impact indicator of alien taxa, we sum all the yearly impact scores of each site of the study region. To correct for sampling effort we divide the yearly impact scores by number of sites in the study region with at least a single occurrence throughout the whole year.\
$$I_i = \frac{\sum{S_i}}{N}$$
$I_i$ is impact score at year $i$.\
$S_i$ is the sum of risk map value, where $S=\{s_1,s_2,...,s_n\}$ and $s_n$ is the site score for site $n$\
$N$ is number of sites occupied through out the study years of the region.\
**Note**: This is the only method incorporated as at now. Other methods will be considered later.\
**Note**: A function `impact_uncertainty()` is being developed to use bootstrap 
method to compute confidence interval of the indicator rather than using `geom_smooth()`
used below.
```{r}
#sum of impact risk map for each year

ggplot(data = impact_value$impact_values) +
  geom_line(aes(y = value, x = year),colour="red",
            stat="identity",
            linewidth=2)+
  geom_smooth(aes(y = value, x = year),linetype=2)+
  labs(
    title = "Impact indicator",
    y = "impact score"
  )+
  theme_minimal() +
  theme(text=element_text(size=14))
```

# Impact indicator per species

We compute the impact indicator per species by summing the impact risk map per species and correct for sampling effort by dividing by $N$.
```{r}
#  impact indicator per species
impact_value$species_values %>%
  rownames_to_column("year") %>%
  mutate(year=as.numeric(year)) %>%
  gather(-year,key = "Alien_species", value = "impact_score") %>%
  ggplot(aes(x = year, y = impact_score)) +
  geom_line(aes(color = Alien_species),linewidth=1.5)+
  theme_minimal() +
  labs(
    title = "sum of species impact",
    y = "impact score"
  )+
  theme(text=element_text(size=14))

```

# Comparing type of indicators

To compare type of impact indicators for a case study, we provide a plot which can
be adapted by a user to compare a set of method.

```{r}
# plot all type of impact indicators
types<-c("precautionary",
         "precautionary cumulative",
         "mean",
         "mean cumulative",
         "cumulative")

all_impact<-data.frame("year"=unique(acacia_cube$cube$data$year))
for(type in types){
  impact_value<-impact_indicator(cube=acacia_cube$cube,
                                 impact_data = eicat_data,
                                 col_impact=NULL,
                                 col_name=NULL,
                                 type = type,
                                 coords=acacia_cube$coords)
  
  all_impact[type]<-impact_value$impact_values$value
}

all_impact %>%
  gather(-year,key = "indicator_type", value = "impact_score") %>% 
  ggplot(aes(x = year, y = impact_score)) +
  geom_line(aes(color = indicator_type),linewidth=1.5)+
  theme_minimal() +
   labs(
    title = "Type of indicators",
    y = "impact score"
  )+
  theme(text=element_text(size=14))
```
